The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2011-10-22 smurf_paper.tex
more edits

The response of the output digitized current, $I$, with respect to a
change in the power $P$ incident on the bolometer, $dI/dP$ depends on
the efficiency of the thermal absorber, how well the TES is biased
into the transition, and the gain of the SQUID amplifiers. Since there
are variations in $T_\mathrm{C}$ across each subarray due to
manufacturing difficulties, considerable effort has gone into
dynamically applying position-dependent biases as well as an over-all
focal plane bias to make as many detectors useable as possible
\citep{holland2011}. In practice, the response of a given detector can
therefore vary somewhat from observation to observation depending on
how the bias and amplification parameters were set up. To track these
changes, at the start of each observation, the current is ramped
through the bolometer heaters such that $dI/dP$ can be measured. Note
that the bolometers may have a slightly different response to heater
power due to conductance as compared to the response to optical power,
so additioanl efficiency corrections must also be obtained. While not
discussed further in this paper, details may be found in
\citet{dempsey2011}. We will refer to measurements of $dI/dP$ across
the focal plane as the \emph{flatfield}. Fig.~\ref{fig:sensitivities}
shows the flatfield for each focal plane.

@@

@@

%-------------------------------------------------
\subsection{Comparison between \scuba\ shared-risk observations and the 2011 re-commissioned instrument}
%-------------------------------------------------

@@

A method that has been employed by teams analyzing BOLOCAM and AzTEC
data to identify and remove low-frequency correlated noise is
Principal Component Analysis
\citep[PCA,][]{laurent2005,scott2008}. While we have not (yet)
included ``PCA cleaning'' as a standard feature of SMURF, we have
performed a preliminary analysis in IDL of some SCUBA-2 data to gain
further insight into the nature of the low-frequency noise. The basic
method is as follows: (i) the covariance matrix is built up for all
pairs $(i,j)$ of the $N$ bolometer time series,
$\langle\mathbf{b}_i(t),\mathbf{b}_j\rangle$; and (ii) an eigenvalue
decomposition is used to identify a new set of statistically
uncorrelated eigenvectors, $\mathbf{\xi}_i(t)$ (i.e. whose covariance
matrix is diagonal), such that each of the bolometer time series is a
linear combination of the eigenvectors, or \emph{components}
i.e. $\mathbf{b}_i(t) = \bar{\mathbf{\xi}} \mathbf{\lambda}_i^T$,
where each row of the matrix $\bar{\mathbf{\xi}}$ is an eigenvector,
and $\mathbf{\lambda}_i^T$ is a column vector containing the
corresponding eigenvalues. For convenience, the $\mathbf{\xi}_i$ are
normalized by their \rms, so that the amplitude of each component is
stored in the eigenvalues. In the earlier analyses mentioned, the
low-frequency noise is assumed to be encapsulated in those components
with the largest eigenvalues. Removing the projection of the time
series along those components has been used successfully to reduce
$1/f$ noise while retaining most of the signal in point-like sources.

@@

Given predictions for the performance of a fully-functioning SCUBA-2,
and the observed properties of the first-installed subarrays during
commissioning in late 2009 as described in the previous section, SMURF
was designed with two primary goals in mind: (i) reduced maps should
be scientifically useful and of ``near publication quality'' with
minimal user interaction; and (ii) the execution time should not
appreciably exceed the amount of time required to conduct
observations. By attaining these goals it would be possible to conduct
real-time reduction at the telescope to offer feedback to observers,
and it would also be possible to maintain a central repository of
(nearly) publishable data products.

There are two general families of existing map-making strategies that
were considered. The first are \emph{maximum likelihood} methods that
were originally pioneered for the analysis of Cosmic Microwave
Background (CMB) maps. Here, the pixels in the map are considered to
be free parameters in a model-fitting problem, and the data are the
raw time-series. By estimating the full noise covariance matrix for
the data, it is then possible to estimate the maximum-likelihood
map. While direct inversion solutions to this problem are untenable,
some recent efforts to optimize these algorithms have enabled the
reduction of smaller data sets in a reasonable amount of time
\citep[e.g.][]{patanchon2008}. While maximum likelihood maps are,
theoretically, the best maps that can be made from a noise point of
view, the required computing resources and execution time would not
have been feasible for the proposed real-time data reduction and data
archiving model.

@@

We express the signal observed by the $i$th bolometer as a function of time,
\begin{equation}
\mathbf{b}_i(t) = g_i[\mathbf{e}_i(t) \mathbf{a}_i(t) + \mathbf{n}_i(t)]
\end{equation}
where $\mathbf{a}(t)$ is the time-varying signal produced by scanning
the telescope across astronomical sources, $\mathbf{e}(t)$ is the
time-varying extinction, which is a function of the telescope
elevation and atmospheric conditions, and $\mathbf{n}_i(t)$ represents
sources of noise. The two terms in square brackets, as written, have
units of power delivered to the detectors (pW). The scale factor $g_i$
converts this effective power to the digitized units recorded on disk
(DAC), and in this formulation is assumed to be constant in time.

In practice, $g_i$ is further factored into two components: $g_i = r_i
d$. The responsivity, $r_i$, gives the change in current as a function
of power fluctuations (A\,pW$^{-1}$). There is then a final (and
constant across all bolometers) conversion factor from current to DAC
units, $d$ (DAC/A).

As for the noise, $\mathbf{n}_i(t)$, we break it down into several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) + \mathbf{n}^\mathrm{f}_i(t) +
  \mathbf{n}^\mathrm{c}(t),
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white (thermal)
noise, $\mathbf{n}^\mathrm{f}_i(t)$ is low-frequency noise that is
\emph{not} correlated from bolometer-to-bolometer, and
$\mathbf{n}^\mathrm{c}(t)$ is a (predominantly low-frequency)
correlated or \emph{common-mode} component. The primary goal of
map-making is to model and remove $\mathbf{n}^\mathrm{f}_i$ and
$\mathbf{n}^\mathrm{c}$ from the time-series. By re-gridding the
remaining data we can then hope to approach the theoretical noise
limit in the map. This procedure is summarized in Fig.~\ref{fig:dimm}.

@@

First, the data files are read from disk, concatenated into contiguous
arrays in memory, and run through a pre-processing stage that performs
several operations including spike and sudden level step removal to
enforce continuity in the time-series. Then, the iterative process
within the dashed box begins. It is generally assumed that the
dominant noise component is the common-mode, which is presumably
dominated by large variations in the fridge temperature and sky noise
(as argued in Section~\ref{sec:pca}). This signal is estimated as the
average signal observed by all of the bolometers at each time step,
and then removed from all of the bolometers. Once this large signal
has been removed, we apply the extinction correction, which is
determined completely from external measurements. Next, residual noise
that differse from bolometer-to-bolometer (primarily low-frequency) is
removed using an FFT-based filter. The resulting ``clean'' time-series
are then re-gridded to produce a map estimate using nearest-neighbour
sampling. Since each map pixel averages many bolometer samples
together, it is fundamentally much less noisy than the bolometer
time-series. We can therefore project the map back into the time
domain, and create relatively noiseless estimates of what each
bolometer would have seen in the absence of noise sources, and remove
that from the time-series. This final step in the iteration gives a
residual signal (with low-frequency and astronomical noise components
removed) that may be used to measured the white noise level in each
detector, and also to estimate an approximate $\chi^2$ for the model
to track convergence. As we will discuss in the next sections, each
signal component estimate is biased slightly from the presence of
signal from other components, so the entire process is
iterated. However, before the next iteration begins, we first undo the
multiplicative factors introduced by the extinction correction, and
also relative gains that may optionally be estimated in the
common-mode estimation phase.

@@

AGG

This is actually only calculated once before the map solution begins
using external opacity monitors (WVM, $\tau_\mathrm{CSO}$,
skydip). Summarize the measurements, order of preference and something
about the calibration for the SCUBA-2 filters.

@@

This model takes the FFT of the bolometer data, and can apply both
high- and low-pass filters, as well as notch filteres, at hard
frequency edges specified by the user. Alternatively, the frequency
edges of the filters may be defined in terms of an \emph{angular
  scale}, but converted into a frequency through knowledge of the mean
telescope slew speed. The time-series may optionally have apodization
applied before the transform to avoid ringing (primarily caused by
wrap-around discontinuities at the ends of the time-series). The
signal that is \emph{removed} from the time-series by this process are
stored in the \texttt{FLT} container array.

@@

Columns share significant magnetic field pickup which is measured by
the dark squids. This model has been included although it makes little
improvement, or injects noise since the magnetic field pickup noise is
generally sub-dominant.

@@

Introduce the topic of model degeneracies here, but defer their
solution to the example sections.

>>>>

The response of the output digitized current, $I$, with respect to a
change in the power $P$ incident on the bolometer, $dI/dP$ depends on
the efficiency of the thermal absorber, how well the TES is biased
into the transition, and the gain of the SQUID amplifiers. Since there
are variations in $T_\mathrm{C}$ across each subarray due to
manufacturing difficulties, considerable effort has gone into
dynamically applying position-dependent biases as well as an over-all
focal plane bias to make as many detectors useable as possible
\citep{holland2011}. In practice, the response of a given detector can
therefore vary somewhat from observation to observation depending on
how the bias and amplification parameters were set up. To track these
responsivity changes, at the start of each observation, the current is
ramped through the bolometer heaters such that $dI/dP$ can be
measured. Note that the bolometers may have a slightly different
response to heater power due to conductance as compared to the
response to optical power, so additioanl efficiency corrections must
also be obtained. While not discussed further in this paper, details
may be found in \citet{dempsey2011}. We will refer to measurements of
$dI/dP$ across the focal plane as the
\emph{flatfield}. Fig.~\ref{fig:sensitivities} shows the flatfield for
each focal plane.

@@

The final conversion of power received by the detector into
astronomical flux units requires two more fundamental corrections: (i)
atmospheric extinction (due to atmospheric water vapour) must be
compensated for; and (ii) the corrected power is related to an
astronomical flux density through the effective area of the detector,
the spectral response of the bolometers (bandpass), the spectrum of
the source, and additional optical efficiency factors -- a \emph{flux
  conversion factor} (FCF). Extinction is measured, in most cases,
using the JCMT line-of-sight water vapour radiometer (WVM) which is
calibrated using skydips and the CSO 225\,GHz tipper. The FCF is
determined empirically using measurements of standard sources. For
further details on these calibration steps, see \citet{dempsey2011}.

@@

%-------------------------------------------------
\subsection{Differences between the S2SRO and 2011 re-commissioned instrument}
%-------------------------------------------------

@@

A method that has been employed by teams analyzing BOLOCAM and AzTEC
data to identify and remove correlated noise is Principal Component
Analysis \citep[PCA,][]{laurent2005,scott2008}. We initially used PCA
to gain further insight into the nature of the \scuba\ noise
properties, but it is now also included it in SMURF as method for
cleaning the data.

The basic method is as follows: (i) a covariance matrix is built up
for all pairs $(i,j)$ of the $N$ bolometer time series,
$\langle\mathbf{b}_i(t),\mathbf{b}_j\rangle$; and (ii) a singular
value decomposition is used to identify a new set of statistically
uncorrelated eigenvectors, $\mathbf{\xi}_i(t)$ (i.e. whose covariance
matrix is diagonal), such that each of the bolometer time series is a
linear combination of the eigenvectors, or \emph{components}
i.e. $\mathbf{b}_i(t) = \bar{\mathbf{\xi}} \mathbf{\lambda}_i^T$,
where each row of the matrix $\bar{\mathbf{\xi}}$ is an eigenvector,
and $\mathbf{\lambda}_i^T$ is a column vector containing the
corresponding eigenvalues. The $\mathbf{\xi}_i$ are normalized by
their \rms, so that the amplitude of each component is stored in the
eigenvalues. In the earlier analyses mentioned, the low-frequency
noise is assumed to be encapsulated in those components with the
largest eigenvalues. Removing the projection of the time series along
those components has been used successfully to reduce $1/f$ noise
while retaining most of the signal in point-like sources. There is an
implicit assumption in this analysis that the noise properties are
stationary.

@@

Given \scuba's data rate, SMURF was designed with two primary goals in
mind: (i) reduced maps should be scientifically useful and of ``near
publication quality'' with minimal user interaction; and (ii) the
execution time should not appreciably exceed the amount of time
required to conduct observations on a high-end multi-tcore desktop
computer. By attaining these goals it would be possible to conduct
real-time reductions at the telescope to offer feedback to observers,
and it would also be possible to maintain a central repository of
(nearly) publishable data products.

As described in the introduction, rather than pursue a slow but
potentially more accurate \emph{maximum likelihood} method that would
require the inversion of a complex linear problem, we instead use an
iterative approach that estimates and removes most of the correlated
(and in particular low-frequency) noise sources in parallel with a
simplified map estimation. We do note, however, that recent advances
in direct inversion methods such as \citet{patanchon2008} may in fact
be tenable for SCUBA-2 data. A future goal of the software team is to
investigate meshing SMURF (perhaps as a cleaning step and initial
condition) for a SANEPIC-like algorithm.

The remainder of this section gives an overview of the model that is
assumed by SMURF for the \scuba\ data, and the typical steps used to
estimate a map. Subsequent sections discuss problems with this model
and additional processing steps and constraints that can be used to
improve the solution in different cases.

@@

We express the signal observed by the $i$th bolometer as a function of time,
\begin{equation}
\mathbf{b}_i(t) = g_i[\mathbf{e}_i(t) \mathbf{a}_i(t) + \mathbf{n}_i(t)]
\end{equation}
where $\mathbf{a}(t)$ is the time-varying signal produced by scanning
the telescope across astronomical sources, $\mathbf{e}(t)$ is the
time-varying extinction, which is a function of the telescope
elevation and atmospheric conditions, and $\mathbf{n}_i(t)$ represents
sources of noise. The two terms in square brackets, as written, have
units of power delivered to the detectors (pW). The scale factor $g_i$
converts this effective power to the digitized units recorded on disk
(DAC) -- the \emph{flatfield} multiplied by a digitization constant --
and in this formulation it is assumed to be constant in time.

We then express the noise, $\mathbf{n}_i(t)$, as the sum of several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) + \mathbf{n}^\mathrm{f}_i(t) +
  \mathbf{n}^\mathrm{c}(t),
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white (thermal)
noise, $\mathbf{n}^\mathrm{f}_i(t)$ is low-frequency noise that is
\emph{not} correlated from bolometer-to-bolometer, and
$\mathbf{n}^\mathrm{c}(t)$ is a (predominantly low-frequency)
correlated or \emph{common-mode} component. The primary goal of
map-making is to model and remove $\mathbf{n}^\mathrm{f}_i$ and
$\mathbf{n}^\mathrm{c}$ from the time-series, while re-gridding the
remaining data we can then hope to approach the theoretical noise
limit in the map. This procedure is summarized in Fig.~\ref{fig:dimm}.

@@

First, the data files are read from disk, concatenated into contiguous
arrays in memory, and run through a pre-processing stage that performs
several operations including spike and sudden level step removal to
enforce continuity in the time-series. Then, the iterative process
within the dashed box begins. It is generally assumed that the
dominant noise component is the common-mode, which is presumably
dominated by large variations in the fridge temperature and sky noise
(as argued in Section~\ref{sec:pca}). This signal is estimated as the
average signal observed by all of the bolometers at each time step,
and then removed from all of the bolometers. Once this large signal
has been removed, we apply the extinction correction, which is
determined completely from external measurements
\citep[see][]{dempsey2011}. Next, residual noise that differs from
bolometer-to-bolometer (primarily low-frequency) is removed, usually
using a high-pass filter. The resulting ``clean'' time-series are then
re-gridded to produce a map estimate using nearest-neighbour
sampling. Since each map pixel is calculated as the average of many
bolometer samples, it is fundamentally much less noisy than the raw
bolometer time-series. We can therefore project the map back into the
time domain (i.e. scanning the array across the current estimate of
the map), and create approximately noiseless estimates of what each
bolometer would have seen in the absence of noise sources, and remove
that from the time-series. This final step in the iteration gives a
residual signal (with low-frequency and astronomical noise components
removed) that may be used to measured the white noise level in each
detector, and also to estimate an approximate $\chi^2$ for the model
to track convergence. As we will discuss in the next sections, each
signal component estimate is biased slightly from the presence of
signal from other components, so the entire process is
iterated. However, before the next iteration begins, we first undo the
multiplicative factors introduced by the extinction correction, and
also relative gains that may optionally be estimated in the
common-mode estimation phase.

@@

The extinction correction is normally derived using the JCMT water
vapour radiometer (WVM), and is not considered to be a free parameter
in the solution. For details on how it is calculated see
\citet{dempsey2011}. Note that its numerical value is calculated only
once, and simply applied as a scale factor in the iterative solution.

@@

This model takes the FFT of the bolometer data, and can apply both
high- and low-pass filters, as well as notch filteres, at hard
frequency edges specified by the user. Alternatively, the frequency
edges of the filters may be defined in terms of an \emph{angular
  scale}, but converted into a frequency through knowledge of the mean
telescope slew speed. The time-series may optionally have apodization
applied before the transform to avoid ringing (primarily caused by
wrap-around discontinuities at the ends of the time-series). Finally,
a whitening filter may also be applied in which a simple $1/f^\alpha +
\mathrm{constant}$ is fit to the power spectrum of each bolometer, and
then the bolometer FFT is multiplied by its inverse (though normalized
to preserve the white noise level).  The signal that is \emph{removed}
from the time-series by this process are stored in the \texttt{FLT}
container array.

@@

Columns share significant magnetic field pickup which is measured by
the dark squids. This model has been included although it makes little
improvement, or injects noise since the magnetic field pickup noise is
generally sub-dominant. Furthermore, not every column that contains
working bolometers also has a working dark squid. An alternative is to
use the \texttt{TMP} model described below.

\subsubsection{\texttt{TMP}: remove a scaled template}

Since the \texttt{DKS} cannot generally be used due to the missing
dark squids, an alternative is to use the telescope pointing as a
proxy for the signal due to magnetic field pickup
(Section~\ref{sec:magpickup}. Generally speaking the azimuth (or its
sine) as a function of time is used as a template that is simply
scaled and removed from each bolometer signal. Some modest
improvements can be achieved with this model.

@@

A fundamental problem faced by the SMURF algorithm, as described, is
that there are degeneracies between model parameters. Thinking of each
pixel in the output map, as well as each sample in the various noise
model components, such as the common-mode and signal removed by the
Fourier filtering, any degeneracy can cause the solution to
diverge. Of particular concern are situations in which the map
diverges.
